---
title: "Activity 3"
subtitle: "Inference: Standard Error Estimation"
toc: true
format: html
---

## Before you begin

Install the required packages (skip any you already have):

```{r}
#| eval: false
install.packages(c("dplyr", "ggplot2", "kableExtra", "broom",
                   "sandwich", "geex"))
```

You also need two files from the workshop in your working directory:

- `data/tpa_ds.rds` — the dataset
- `R/metadata.R` — variable labels and factor levels

Download them from the workshop website and place them in your working directory before running this file.

---

In Activity 3, we estimate standard errors for the IPT-weighted (ATE) and SMR-weighted (ATT) risk differences using three approaches: the robust sandwich variance, the bootstrap, and M-estimation.

Standard error estimation for weighted estimators is non-trivial. The weights are themselves estimated quantities, and their uncertainty should be propagated through to the final estimate. The three methods handle this very differently.

# Setup

Run this block first to load packages, data, and reproduce the preprocessing and weight estimation from Activities 1 and 2.

```{r}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(broom)
library(sandwich)
library(geex)
```

```{r}
source("R/metadata.R")
tpa_ds <- readRDS("data/tpa_ds.rds")

tpa_ds$age_5f <- cut(tpa_ds$age, breaks = seq(14, 105, by = 5), right = FALSE)

exposure    <- "tpa"
outcome     <- "death"
confounders <- c("age", "sex", "living", "comorbid", "cardiac", "prevstroke",
                 "htn", "hyperchol", "afib", "aphasia", "paresis_3f", "rankpre", "icu")

tpa_ds_cc <- tpa_ds |>
  dplyr::filter(dplyr::if_all(dplyr::all_of(c(exposure, outcome, confounders)), ~!is.na(.x)))

## Propensity scores and weights
expform <- sprintf("tpa ~ %s", paste(confounders, collapse = " + "))
ps_model <- glm(as.formula(expform), data = tpa_ds_cc, family = binomial(link = "logit"))
tpa_ds_cc$ps <- predict(ps_model, type = "response")

tpa_ds_cc$ipt <- with(tpa_ds_cc, tpa / ps + (1 - tpa) / (1 - ps))
tpa_ds_cc$smr <- with(tpa_ds_cc, tpa + (1 - tpa) * ps / (1 - ps))

## Weighted outcome models
ate_rd_model <- lm(death ~ tpa, data = tpa_ds_cc, weights = ipt)
att_rd_model <- lm(death ~ tpa, data = tpa_ds_cc, weights = smr)

cat("ATE RD:", round(coef(ate_rd_model)["tpa"], 4), "\n")
cat("ATT RD:", round(coef(att_rd_model)["tpa"], 4), "\n")
```

---

# Method 1: Robust sandwich variance

The sandwich (Huber-White) variance estimator is a closed-form, fast approach. It accounts for heteroskedasticity in the weighted regression, but **treats the estimated weights as fixed and known** — it does not account for the uncertainty from estimating the propensity score.

For the ATE, this means the sandwich SE is generally **conservative** (too large). For the ATT (which uses inverse-odds weights), it can be either conservative or anti-conservative.

$$\hat{V}_{\text{sandwich}} = (\hat{A})^{-1} \hat{B} \left[(\hat{A})^{-1}\right]^T$$

```{r}
ate_se_robust <- sqrt(sandwich::sandwich(ate_rd_model)[2, 2])
att_se_robust <- sqrt(sandwich::sandwich(att_rd_model)[2, 2])

cat("Robust SE (ATE RD):", round(ate_se_robust, 4), "\n")
cat("Robust SE (ATT RD):", round(att_se_robust, 4), "\n")
```

---

# Method 2: Bootstrap

The bootstrap resamples the *entire dataset* with replacement and re-estimates everything from scratch on each resample: the propensity score model, the weights, and the outcome model. The standard deviation of the bootstrap estimates is the bootstrap SE.

This correctly captures all sources of variability, including uncertainty from estimating the propensity score. The drawback is computational cost.

```{r}
set.seed(24601)
B <- 500

boot_results <- matrix(nrow = B, ncol = 2)  # col 1 = ATE RD, col 2 = ATT RD

for (i in seq_len(B)) {
  boot_idx    <- sample(nrow(tpa_ds_cc), replace = TRUE)
  boot_sample <- tpa_ds_cc[boot_idx, ]

  ## Re-estimate propensity score on bootstrap sample
  ps_boot          <- glm(as.formula(expform), data = boot_sample,
                          family = binomial(link = "logit"))
  boot_sample$ps   <- predict(ps_boot, type = "response")

  ## Re-calculate weights
  boot_sample$ipt  <- with(boot_sample, tpa / ps + (1 - tpa) / (1 - ps))
  boot_sample$smr  <- with(boot_sample, tpa + (1 - tpa) * ps / (1 - ps))

  ## Re-estimate weighted outcome models
  boot_results[i, 1] <- lm(death ~ tpa, data = boot_sample,
                            weights = ipt)$coefficients["tpa"]
  boot_results[i, 2] <- lm(death ~ tpa, data = boot_sample,
                            weights = smr)$coefficients["tpa"]
}

boot_se <- apply(boot_results, 2, sd)
names(boot_se) <- c("ATE", "ATT")

cat("Bootstrap SE (ATE RD):", round(boot_se["ATE"], 4), "\n")
cat("Bootstrap SE (ATT RD):", round(boot_se["ATT"], 4), "\n")
```

You can also visualize the bootstrap distribution to check that it looks approximately normal, which is a requirement for valid Wald-type confidence intervals.

```{r}
par(mfrow = c(1, 2))
hist(boot_results[, 1], main = "Bootstrap: ATE RD", xlab = "Estimate", breaks = 30)
hist(boot_results[, 2], main = "Bootstrap: ATT RD", xlab = "Estimate", breaks = 30)
```

---

# Method 3: M-estimation

M-estimation propagates uncertainty from *all* estimation steps — propensity score model and outcome model — simultaneously through a system of **stacked estimating equations**. It is computationally efficient (no resampling) and asymptotically valid.

An M-estimator $\hat{\boldsymbol{\theta}}$ solves:

$$\sum_{i=1}^n \psi\left(\mathbf{Y}_i, \hat{\boldsymbol{\theta}}\right) = \mathbf{0}$$

where $\psi$ stacks the score equations from the propensity score model with the estimating equations for the weighted means and risk differences.

The variance of $\hat{\boldsymbol{\theta}}$ is estimated by the empirical sandwich variance:

$$\hat{V} = \hat{A}^{-1} \hat{B} \left(\hat{A}^{-1}\right)^T, \quad \hat{A} = \frac{1}{n}\sum_i \left\{-\psi'\left(Y_i, \hat{\boldsymbol{\theta}}\right)\right\}, \quad \hat{B} = \frac{1}{n}\sum_i \psi\left(Y_i, \hat{\boldsymbol{\theta}}\right)\psi\left(Y_i, \hat{\boldsymbol{\theta}}\right)^T$$

The key benefit: because the propensity score equations are *part of the stack*, their uncertainty is automatically accounted for in the final SEs.

## Set up the data matrix

We first create a design matrix with indicator variables for all multi-level covariates. This is needed because the propensity score estimating equations require a numeric design matrix.

```{r}
mest_data <- tpa_ds_cc |>
  dplyr::mutate(
    paresis_mild     = as.integer(paresis_3f == "Mild"),
    paresis_severe   = as.integer(paresis_3f == "Severe"),
    living_home      = as.integer(living == "At home"),
    rankpre_nosig    = as.integer(rankpre == "No significant disability"),
    rankpre_slight   = as.integer(rankpre == "Slight disability"),
    rankpre_moderate = as.integer(rankpre == "Moderate disability"),
    rankpre_severe   = as.integer(rankpre == "Severe disability")
  ) |>
  dplyr::select(tpa, death, sex, age, comorbid, cardiac, prevstroke,
                htn, hyperchol, afib, aphasia, icu,
                paresis_mild, paresis_severe, living_home,
                starts_with("rankpre_"))

cat("Dimensions of M-estimation data:", nrow(mest_data), "x", ncol(mest_data), "\n")
cat("Number of PS parameters (including intercept):", ncol(mest_data) - 1, "\n")
```

## Define the stacked estimating function

The estimating function stacks:

1. Propensity score model (logistic regression score equations — one per PS parameter)

2. IPT-weighted group means (control and treated)

3. ATE risk difference definition: $\theta_{\text{ATE}} = \mu_1^{\text{IPT}} - \mu_0^{\text{IPT}}$

4. SMR-weighted group means (control and treated)

5. ATT risk difference definition: $\theta_{\text{ATT}} = \mu_1^{\text{SMR}} - \mu_0^{\text{SMR}}$

```{r}
setup_ef <- function(data) {
  a <- data$tpa
  y <- data$death
  x <- model.matrix(~ ., data = data[, !(names(data) %in% c("tpa", "death"))])
  p_ps <- ncol(x)

  function(theta) {
    theta_ps       <- theta[1:p_ps]
    theta_ipt_mu   <- theta[(p_ps + 1):(p_ps + 2)]  # [mu0_ipt, mu1_ipt]
    theta_ipt_ate  <- theta[p_ps + 3]
    theta_smr_mu   <- theta[(p_ps + 4):(p_ps + 5)]  # [mu0_smr, mu1_smr]
    theta_smr_att  <- theta[p_ps + 6]

    ## Propensity score: logistic regression score equations
    lp     <- x %*% theta_ps
    pscore <- plogis(lp)
    ef_ps  <- as.vector(t(x) %*% (a - pscore))

    ## IPT and SMR weights
    ipt <- a / pscore + (1 - a) / (1 - pscore)
    smr <- a + (1 - a) * pscore / (1 - pscore)

    ## Estimating equations for weighted group means
    ef_ipt_mu0  <- ((1 - a) * ipt) * (y - theta_ipt_mu[1])
    ef_ipt_mu1  <- (a * ipt) * (y - theta_ipt_mu[2])
    ef_smr_mu0  <- ((1 - a) * smr) * (y - theta_smr_mu[1])
    ef_smr_mu1  <- (a * smr) * (y - theta_smr_mu[2])

    ## Estimating equations for ATE and ATT risk differences
    ef_ate <- theta_ipt_mu[2] - theta_ipt_mu[1] - theta_ipt_ate
    ef_att <- theta_smr_mu[2] - theta_smr_mu[1] - theta_smr_att

    c(ef_ps,
      ef_ipt_mu0, ef_ipt_mu1, ef_ate,
      ef_smr_mu0, ef_smr_mu1, ef_att)
  }
}
```

## Solve and compute the sandwich variance

`geex::m_estimate()` handles both steps: it finds $\hat{\boldsymbol{\theta}}$ by solving the stacked estimating equations, then computes the sandwich variance automatically. We supply starting values and pass `setup_ef` directly as the estimating function — `geex` calls it once per observation.

```{r}
p_ps    <- ncol(model.matrix(~ ., data = mest_data[, !(names(mest_data) %in% c("tpa","death"))]))
mu1_est <- mean(mest_data$death[mest_data$tpa == 1])
mu0_est <- mean(mest_data$death[mest_data$tpa == 0])
start   <- c(rep(0, p_ps), mu0_est, mu1_est, mu1_est - mu0_est,
             mu0_est, mu1_est, mu1_est - mu0_est)

geex_result <- m_estimate(
  estFUN       = setup_ef,
  data         = mest_data,
  root_control = setup_root_control(start = start)
)

theta_hat <- roots(geex_result)

idx_ate <- p_ps + 3
idx_att <- p_ps + 6

mest_se <- c(
  ATE = sqrt(vcov(geex_result)[idx_ate, idx_ate]),
  ATT = sqrt(vcov(geex_result)[idx_att, idx_att])
)

cat("M-estimation SE (ATE RD):", round(mest_se["ATE"], 4), "\n")
cat("M-estimation SE (ATT RD):", round(mest_se["ATT"], 4), "\n")

## Check: M-estimation estimates should match our earlier weighted estimates
cat("\nM-estimation ATE RD estimate:", round(theta_hat[idx_ate], 4), "\n")
cat("Weighted regression ATE RD:  ", round(coef(ate_rd_model)["tpa"], 4), "\n")
```

---

# Comparing all three methods

```{r}
tibble::tibble(
  Estimand  = c("ATE RD", "ATT RD"),
  Estimate  = round(c(coef(ate_rd_model)["tpa"],
                      coef(att_rd_model)["tpa"]), 4),
  `Robust SE`   = round(c(ate_se_robust, att_se_robust), 4),
  `Bootstrap SE` = round(boot_se, 4),
  `M-estimation SE` = round(mest_se, 4)
) |>
  kable()
```

## Discussion

- **Why do the estimates match across methods?** All three methods estimate the same quantity; they differ only in how they compute uncertainty.

- **Robust sandwich SE**: Fast and closed-form, but treats weights as fixed. Generally conservative for the ATE (overestimates SE). For the ATT, it can go either way — see Reifeis & Hudgens (2022) *AJE*.

- **Bootstrap SE**: Resamples the entire data-generating process, capturing all variability including from estimating the propensity score. Computationally expensive but valid.

- **M-estimation SE**: Analytically propagates uncertainty from the propensity score model to the outcome, yielding consistent SEs. Efficient (no resampling), but relies on large-sample asymptotics.
