---
title: Data
toc: true
---

# Data Source

- Originates from the Westphalian Stroke Register: regional data bank in the northwest of Germany

- On a voluntary basis, 24 neurological, 13 internal medicine and 5 geriatric medicine departments in 42 hospitals participated in the project in 2000 and 2001

- Included patients treated for a transient ischemic attack, ischemic stroke or intracerebral hemorrhage

- Sociodemographic characteristics, cerebrovascular risk factors, comorbidities, stroke type and severity, as well as details regarding the participating centers, the mode of admission, diagnostic and therapeutic procedures and complications documented

- The standardized data assessment forms were evaluated centrally at the Institute of Epidemiology and Social Medicine at the University of Muenster, Germany

- In total, 12,410 patients with stroke symptoms were entered into the registry until 2001: 8,208 were ischemic, 2,794 transient ischemic attacks, 793 primary intracerebral hemorrhages and 615 unknown

- For this clinical example, only ischemic stroke patients were considered. Furthermore, ischemic stroke patients were excluded from centers that did not perform t-PA lysis throughout the time of observation (2000-2001)

- A total of 6,269 patients were included in the original example. More details of the original sample can be found: Kurth T et al. Results of multivariable logistic regression, propensity matching, propensity adjustment, and propensity-based weighting under conditions of nonuniform effect. Am J Epidemiol. 2006 Feb 1;163(3):262--70.

- The data set we use in this lab is a **randomly modified dataset**. Therefore, some of the association pattern behaves differently than clinically expected. The total number of observations in this dataset is set to 10,000.

- This data set is for your **personal use only**. Please do not share it outside of class.

# Data Dictionary

```{r, echo = F}
library(kableExtra)
source("R/metadata.R")
purrr::imap_dfr(
    varlabels,
    .f =   function(x, y){
      dplyr::tibble(
        name = y,
        label = x$lbl,
        type = x$type,
        values = if (type == "factor") {
          paste0(x$lvl, " = ", names(x$lvl),collapse = "<br>")
        } else if (type == "binary") {
          "1 = Yes <br> 0 = No"
        } else if (type == "numeric") {
          paste0(x$lvl) 
        }
      )
    }
  ) |> 
  dplyr::distinct()-> df
```

```{r, echo = F}
#| results: asis

df|> 
  dplyr::arrange(name) |> 
  kableExtra::kable(escape = FALSE,
                    format = "html",
                   # longtable = TRUE,
                    col.names = c("Variable Name", "Variable Label", "Variable Type", "Values")) |> 
  kable_styling()

```

